---
title: "Grade A Wine EDA"
output: 
  github_document: default
date: "2024-10-28"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 7, fig.height = 5)
```




Importing necessary libraries, importing data, and readying data for analysis
```{r, echo=FALSE}
library(corrplot)
library(Hmisc)
library(ggplot2)
library(psych)
library(GGally)
library(vioplot)
library(DescTools)
library(leaps)
library(tidyverse)
library(caret)
library(e1071)
library(rattle)
library(dplyr)
library(rpart)
library(kknn)
library(stats)
library(factoextra)
library(MASS)
library(car)
library(xgboost)
library(tidyverse)
library(data.table)
library(skimr)
library(randomForest)
library(ROSE)
library(tuneRanger)
library(VSURF)
library(foreach)
library(doParallel)

```


```{r}

setwd("~/Downloads")
redwine <- read.csv("winequality-red.csv")
whitewine <- read.csv("winequality-white.csv")

cat("Creating variable Names Red Wine")
redwine_seperated <- str_split_fixed(redwine$fixed.acidity.volatile.acidity.citric.acid.residual.sugar.chlorides.free.sulfur.dioxide.total.sulfur.dioxide.density.pH.sulphates.alcohol.quality, ';', 12)

redwine_seperated <- data.frame(redwine_seperated)

cat("Creating variable Names White Wine")
whitewine_seperated <- str_split_fixed(whitewine$fixed.acidity.volatile.acidity.citric.acid.residual.sugar.chlorides.free.sulfur.dioxide.total.sulfur.dioxide.density.pH.sulphates.alcohol.quality, ';', 12)

whitewine_seperated <- data.frame(whitewine_seperated)



redwine_seperated <- redwine_seperated %>% 
  rename(fixed_acidity = 'X1')
redwine_seperated <- redwine_seperated %>% 
  rename(volatile_acidity = 'X2')
redwine_seperated <- redwine_seperated %>% 
  rename(citric_acid = 'X3')
redwine_seperated <- redwine_seperated %>% 
  rename(residual_sugar = 'X4')
redwine_seperated <- redwine_seperated %>% 
  rename(chlorides = 'X5')
redwine_seperated <- redwine_seperated %>% 
  rename(free_sulfur_dioxide = 'X6')
redwine_seperated <- redwine_seperated %>% 
  rename(total_sulfur_dioxide = 'X7')
redwine_seperated <- redwine_seperated %>% 
  rename(density = 'X8')
redwine_seperated <- redwine_seperated %>% 
  rename(pH = 'X9')
redwine_seperated <- redwine_seperated %>% 
  rename(sulphates = 'X10')
redwine_seperated <- redwine_seperated %>% 
  rename(alcohol = 'X11')
redwine_seperated <- redwine_seperated %>% 
  rename(quality = 'X12')

whitewine_seperated <- str_split_fixed(whitewine$fixed.acidity.volatile.acidity.citric.acid.residual.sugar.chlorides.free.sulfur.dioxide.total.sulfur.dioxide.density.pH.sulphates.alcohol.quality, ';', 12)

whitewine_seperated <- data.frame(whitewine_seperated)

whitewine_seperated <- whitewine_seperated %>% 
  rename(fixed_acidity = 'X1')
whitewine_seperated <- whitewine_seperated %>% 
  rename(volatile_acidity = 'X2')
whitewine_seperated <- whitewine_seperated %>% 
  rename(citric_acid = 'X3')
whitewine_seperated <- whitewine_seperated %>% 
  rename(residual_sugar = 'X4')
whitewine_seperated <- whitewine_seperated %>% 
  rename(chlorides = 'X5')
whitewine_seperated <- whitewine_seperated %>% 
  rename(free_sulfur_dioxide = 'X6')
whitewine_seperated <- whitewine_seperated %>% 
  rename(total_sulfur_dioxide = 'X7')
whitewine_seperated <- whitewine_seperated %>% 
  rename(density = 'X8')
whitewine_seperated <- whitewine_seperated %>% 
  rename(pH = 'X9')
whitewine_seperated <- whitewine_seperated %>% 
  rename(sulphates = 'X10')
whitewine_seperated <- whitewine_seperated %>% 
  rename(alcohol = 'X11')
whitewine_seperated <- whitewine_seperated %>% 
  rename(quality = 'X12')


redwine_seperated <- apply(redwine_seperated,2,as.numeric)
whitewine_seperated <- apply(whitewine_seperated,2,as.numeric)

redwine_seperated <- data.frame(redwine_seperated)
whitewine_seperated <- data.frame(whitewine_seperated)



redwine_seperated$type <- 'red'
whitewine_seperated$type <- 'white'

redwine_seperated$type <- as.factor(redwine_seperated$type)
whitewine_seperated$type <- as.factor(whitewine_seperated$type)

wine <- full_join(redwine_seperated,whitewine_seperated)

```


Exploratory Data Analysis Getting Descriptive Statistics for Red and White Wine
```{r}
#Getting descriptive stats for Red Wine
skim(redwine_seperated)

#Basic Stats
summary(redwine_seperated)



#Getting descriptive stats for White Wine
skim(whitewine_seperated)


#Basic Stats
summary(whitewine_seperated)

```



EDA Continued: Exploring Data by viewing distributions, 
    and frequency of outliers for each variable
```{r, echo=FALSE, fig.width=7, fig.height=5}

# Box plot Fixed Acidity grouped by Wine Type
ggplot(wine, aes(x = type, y = fixed_acidity)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(x = "Wine Type", y = "Fixed Acidity", 
       title = "Box Plot of Fixed Acidity by Wine Type")

# Histogram of Fixed Acidity with bars colored by Wine Type
ggplot(wine, aes(x = fixed_acidity, fill = type,colour = type)) +
  geom_histogram(alpha = 0.5, position = "identity") +
  labs(x = "Wine Type", y = "Count", 
       title = "Histogram of Fixed Acidity by Wine Type")



# Box plot Volatile Acidity grouped by Wine Type
ggplot(wine, aes(x = type, y = volatile_acidity)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(x = "Wine Type", y = "Volatile Acidity", 
       title = "Box Plot of Volatile Acidity by Wine Type")

# Histogram of Volatile Acidity  with bars colored by Wine Type
ggplot(wine, aes(x = volatile_acidity, fill = type,colour = type)) +
  geom_histogram(alpha = 0.5, position = "identity") +
  labs(x = "Wine Type", y = "Count", 
       title = "Histogram of Volatile Acidity by Wine Type")



# Box plot Citric Acid grouped by Wine Type
ggplot(wine, aes(x = type, y = citric_acid)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(x = "Wine Type", y = "Citric Acid", 
       title = "Box Plot of Citric Acid by Wine Type")

# Histogram of Citric Acid with bars colored by Wine Type
ggplot(wine, aes(x = citric_acid, fill = type,colour = type)) +
  geom_histogram(alpha = 0.5, position = "identity") +
  labs(x = "Wine Type", y = "Count", 
       title = "Histogram of Citric Acid by Wine Type")



# Box plot Residual Sugar grouped by Wine Type
ggplot(wine, aes(x = type, y = residual_sugar)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(x = "Wine Type", y = "Residual Sugar", 
       title = "Box Plot of Residual Sugar by Wine Type")

# Histogram of Residual Sugar with bars colored by Wine Type
ggplot(wine, aes(x = residual_sugar, fill = type,colour = type)) +
  geom_histogram(alpha = 0.5, position = "identity") +
  labs(x = "Wine Type", y = "Count", 
       title = "Histogram of Residual Sugar by Wine Type")



# Box plot Chlorides grouped by Wine Type
ggplot(wine, aes(x = type, y = chlorides)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(x = "Wine Type", y = "Chlorides", 
       title = "Box Plot of Chlorides by Wine Type")

# Histogram of Chlorides with bars colored by Wine Type
ggplot(wine, aes(x = chlorides, fill = type,colour = type)) +
  geom_histogram(alpha = 0.5, position = "identity") +
  labs(x = "Wine Type", y = "Count", 
       title = "Histogram of Chlorides by Wine Type")



# Box plot Free Sulfur Dioxide grouped by Wine Type
ggplot(wine, aes(x = type, y = free_sulfur_dioxide)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(x = "Wine Type", y = "Free Sulfur Dioxide", 
       title = "Box Plot of Free Sulfur Dioxide by Wine Type")

# Histogram of Free Sulfur Dioxide with bars colored by Wine Type
ggplot(wine, aes(x = free_sulfur_dioxide, fill = type,colour = type)) +
  geom_histogram(alpha = 0.5, position = "identity") +
  labs(x = "Wine Type", y = "Count", 
       title = "Histogram of Free Sulfur Dioxide by Wine Type")



# Box plot Total Sulfur Dioxide grouped by Wine Type
ggplot(wine, aes(x = type, y = total_sulfur_dioxide)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(x = "Wine Type", y = "Total Sulfur Dioxide", 
       title = "Box Plot of Total Sulfur Dioxide by Wine Type")

# Histogram of Total Sulfur Dioxide with bars colored by Wine Type
ggplot(wine, aes(x = total_sulfur_dioxide, fill = type,colour = type)) +
  geom_histogram(alpha = 0.5, position = "identity") +
  labs(x = "Wine Type", y = "Count", 
       title = "Histogram of Total Sulfur Dioxide by Wine Type")



# Box plot Density grouped by Wine Type
ggplot(wine, aes(x = type, y = density)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(x = "Wine Type", y = "Density", 
       title = "Box Plot of Density by Wine Type")

# Histogram of Density with bars colored by Wine Type
ggplot(wine, aes(x = density, fill = type,colour = type)) +
  geom_histogram(alpha = 0.5, position = "identity") +
  labs(x = "Wine Type", y = "Count", 
       title = "Histogram of Density by Wine Type")



# Box plot pH grouped by Wine Type
ggplot(wine, aes(x = type, y = pH)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(x = "Wine Type", y = "pH", title = "Box Plot of pH by Wine Type")

# Histogram of pH with bars colored by Wine Type
ggplot(wine, aes(x = pH, fill = type,colour = type)) +
  geom_histogram(alpha = 0.5, position = "identity") +
  labs(x = "Wine Type", y = "Count", title = "Histogram of pH by Wine Type")



# Box plot Sulfates grouped by Wine Type
ggplot(wine, aes(x = type, y = sulphates)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(x = "Wine Type", y = "Sulfates", 
       title = "Box Plot of Sulfates by Wine Type")

# Histogram of Sulfates with bars colored by Wine Type
ggplot(wine, aes(x = sulphates, fill = type,colour = type)) +
  geom_histogram(alpha = 0.5, position = "identity") +
  labs(x = "Wine Type", y = "Count", 
       title = "Histogram of Sulfates by Wine Type")



# Box plot Alcohol grouped by Wine Type
ggplot(wine, aes(x = type, y = alcohol)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(x = "Wine Type", y = "Alcohol", 
       title = "Box Plot of Alcohol by Wine Type")

# Histogram of Alcohol with bars colored by Wine Type
ggplot(wine, aes(x = alcohol, fill = type,colour = type)) +
  geom_histogram(alpha = 0.5, position = "identity") +
  labs(x = "Wine Type", y = "Count", 
       title = "Histogram of Alcohol by Wine Type")



# Box plot Quality grouped by Wine Type
ggplot(wine, aes(x = type, y = quality)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(x = "Wine Type", y = "Quality", 
       title = "Box Plot of Quality by Wine Type")

# CountPlot of Quality with bars colored by Wine Type
ggplot(wine, aes(x = quality, fill = type,colour = type)) +
  geom_histogram(alpha = 0.5, position = "identity") +
  labs(x = "Wine Type", y = "Count", 
       title = "CountPlot of Quality by Wine Type")


cat("Examining the CountPlot of Wine Type, the purpose of this investigation
    since less than 25% of wines in both red and white wines have high wine 
    ratings greater than 7, Grade A wine is defined as those with an rating of 7 
    or greater. The purpose of this investigation is to predict Grade A wine 
    and be able to define characteristics of Grade A wine")

redwine_seperated$quality <- ifelse(redwine_seperated$quality >= 7,1,0)

whitewine_seperated$quality <- ifelse(whitewine_seperated$quality >= 7,1,0)  

```


EDA Continued:
Checking Correlation of all Variables with correlation plot
```{r, echo=FALSE, fig.width=7, fig.height=5}
#Correlation Plot of all red wine variables  
corrplot(cor(redwine_seperated[1:12]))
title("Correlation Plot of all Red Wine Variables")


#Correlation Plot of all white wine variables 
corrplot(cor(whitewine_seperated[1:12]))
title("Correlation Plot of all White Wine Variables")



```



Since summary statistics showed that variables were on different scales 
    and there was a lot of difference observed between variables must apply 
    scaling method to normalize data.
```{r}
#Applying Min-max scaling 
min_max_scaling_white <- preProcess(whitewine_seperated[1:11], method = "range")
white_wine_scaled <- predict(min_max_scaling_white,whitewine_seperated)
white_wine_scaled$quality <- as.factor(white_wine_scaled$quality)

min_max_scaling_red <- preProcess(redwine_seperated[1:11], method = "range")
red_wine_scaled <- predict(min_max_scaling_red,redwine_seperated)
red_wine_scaled$quality <- as.factor(red_wine_scaled$quality)


redwine_randomforest_columns12 <- c("fixed_acidity", "volatile_acidity",
                                    "citric_acid","residual_sugar","chlorides", 
                                    "free_sulfur_dioxide","total_sulfur_dioxide",
                            "density","pH","sulphates", "alcohol", "quality")
whitewine_randomforest_columns12 <- c("fixed_acidity", "volatile_acidity",
                                      "citric_acid","residual_sugar","chlorides", 
                                      "free_sulfur_dioxide","total_sulfur_dioxide",
                            "density","pH","sulphates", "alcohol", "quality")

red_wine_scaled <- red_wine_scaled[, redwine_randomforest_columns12, 
                                   drop = FALSE ]
white_wine_scaled <- white_wine_scaled[, whitewine_randomforest_columns12, 
                                       drop = FALSE ]




```

Since summary statistics showed that variables were on different scales 
    and there was a lot of difference observed between variables must apply 
    scaling method to normalize data.
    
    

